Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-07-23 manuscript/climate_sensitivity_draft.Rmd
fixed refs

Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought resistance, defined as the ratio of annual growth in a drought year to that which would be expected in the absence of drought based on previous years' growth. 
We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from the three strongest droughts over a 60-year period (1950 - 2009), species trait measurements, and census and microenvironmental data from a large forest dynamics plot in Virginia, USA. 
First, we focus on how tree size, alone and in its interaction with microenvironmental gradients, influences drought resistance. 
We examine the contemporary relationship between tree height and microenvironment, including growing season meteorological conditions and crown exposure. 
We then test whether, consistent with most forests globally, larger-diameter, taller trees tend to have lower drought resistance in this forest, which is in a region (eastern North America) represented by only two studies in the global review of Bennett et al. (2015).
We also test for an influence of potential access to available soil water, which should be greater for larger trees in dry but not in perpetually wet microsites. 
Finally, we focus on the role of species' traits, testing the hypothesis that species’ traits-–particularly leaf leaf drought tolerance traits-–predict drought resistance. 
We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species and that it is correlated with wood density--either positively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011] and positively correlated with $LMA$.
We further test predictions that species with low $PLA_{dry}$ have higher drought resistance, and that species whose leaves lose turgor lower water potentials (more negative $\pi_{tlp}$) have higher resistance.

@@

Belowground, taller trees would tend to have larger root systems [@enquist_global_2002], but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought.
While tree size can correlate with the depth of water extraction [@brum_hydrological_2018], the linkage is not consistent. Shorter trees can vary broadly in the depth of water uptake [@stahl_depth_2013], and larger trees may allocate more to abundant shallow roots that are beneficial for taking up water from rainstorms [@meinzer_partitioning_1999].
Moreover, reliance on deep soil-water resources can actually prove a liability during severe and prolonged drought, as these can experience more intense water scarcity relative to non-drought conditions [@chitra-tarak_roots_2018].
In any case, the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
This may be because moister habitats would tend to support species and individuals with more mesophytic traits [@bartlett_drought_2016; @mencuccini_ecological_2003; @medeiros_extensive_2019], potentially growing to greater heights (e.g., @detto_hydrological_2013), and these are then more vulnerable when drought hits.
The observed height-sensitivity of $Rt$, together with the lack of conferred advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests. 
Again, additional research comparing drought responses across forests with different tree heights and water availability would be valuable for disentangling the relative importance of above- and belowground mechanisms across trees of different size.

>>>>

Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought resistance, defined as the ratio of annual growth in a drought year to that which would be expected in the absence of drought based on previous years' growth. 
We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from the three strongest droughts over a 60-year period (1950 - 2009), species trait measurements, and census and microenvironmental data from a large forest dynamics plot in Virginia, USA. 
First, we focus on how tree size, alone and in its interaction with microenvironmental gradients, influences drought resistance. 
We examine the contemporary relationship between tree height and microenvironment, including growing season meteorological conditions and crown exposure. 
We then test whether, consistent with most forests globally, larger-diameter, taller trees tend to have lower drought resistance in this forest, which is in a region (eastern North America) represented by only two studies in the global review of Bennett et al. (2015).
We also test for an influence of potential access to available soil water, which should be greater for larger trees in dry but not in perpetually wet microsites. 
Finally, we focus on the role of species' traits, testing the hypothesis that species’ traits-–particularly leaf drought tolerance traits-–predict drought resistance. 
We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species and that it is correlated with wood density--either positively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011] and positively correlated with $LMA$.
We further test predictions that species with low $PLA_{dry}$ have higher drought resistance, and that species whose leaves lose turgor lower water potentials (more negative $\pi_{tlp}$) have higher resistance.

@@

Belowground, taller trees would tend to have larger root systems [@enquist_global_2002], but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought.
While tree size can correlate with the depth of water extraction [@brum_hydrological_2019], the linkage is not consistent. Shorter trees can vary broadly in the depth of water uptake [@stahl_depth_2013], and larger trees may allocate more to abundant shallow roots that are beneficial for taking up water from rainstorms [@meinzer_partitioning_1999].
Moreover, reliance on deep soil-water resources can actually prove a liability during severe and prolonged drought, as these can experience more intense water scarcity relative to non-drought conditions [@chitratarak_roots_2018].
In any case, the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
This may be because moister habitats would tend to support species and individuals with more mesophytic traits [@bartlett_drought_2016; @mencuccini_ecological_2003; @medeiros_extensive_2019], potentially growing to greater heights (e.g., @detto_hydrological_2013), and these are then more vulnerable when drought hits.
The observed height-sensitivity of $Rt$, together with the lack of conferred advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests. 
Again, additional research comparing drought responses across forests with different tree heights and water availability would be valuable for disentangling the relative importance of above- and belowground mechanisms across trees of different size.
